---
title: "Activity3_ME"
author: "Suyog Timalsina"
output: html_document
date: "2025-02-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Getting Started With R/RStudio and Data Cleaning Basic

### All the topics we gonna learn in this Activity
* [1. Using `TinyTex`]
* [2. Vectors] 
* [2.1 Vector with numerical values in a sequence]
* [2.2  Generates a vector of random numbers following a normal distribution using rnorm() function]
* [Matrices ]
* [3.1  Operations on Matrices in R]

## 1. Using `TinyTex`

## 2. Vectors

-   to combine the list of items to a vector, we use the `c()` function and separate the items by a comma.
-   Vector of numerical values

```{r}
x <- c(-6,-3,-4,9,10,11)
x

# to access one of the element
x[2]

# All but the second element
x[-2]

# accessing first and fourth element
x[c(1,4)]

#length 
length(x)
```

### 2.1 Vector with numerical values in a sequence

```{r}
# A sequence of values from 1 to 12
y <- 1:12
y
```

```{r}
# Vector with numerical decimals in a sequence
y <- 1.2:5.2
y
```

-   **Sort a Vector using the sort() function**

```{r}
sortX <- c(95,65,53,-58,569,-4886)

sorted_x_desc <- sort(sortX, decreasing = TRUE )  #Descending order
print(sorted_x_desc)

```

-   **Sort a Vector which has value alphabetical**

```{r}
color <- c("apple", "orange", "lime","red","blue")
color
```

-   **Change item**

```{r}
color[1] <- "black"
color
```

-   **to repeat vectors, use the `rep()` function**

```{r}
# Repeat each value three-time
repeat_three_time <- rep(c(2,4,6), each = 3)

repeat_three_time
```

```{r}
# Repeat each value independently
repeat_inde <- rep(c(2,4,6), times = c(4,2,1))

repeat_inde
```

-   **to make bigger or smaller steps in a sequence, `use the seq()` function**

```{r}
w <- seq(from = 0 , to = 200, by= 20)
w
```

-   **Creating an empty vector**

```{r}
z <- vector(length = 5)
z
```

```{r}
z[1] <- 110
z[4] <- 15
z[5] <- x[4] + 5
z
```

-   **Operation on Vectors**

```{r}
x <- c(5, 0, 3, 4, 7, 20)
k <- c(6, 8, 1, 5, 2, 5)

# addition
d <- x + k
d
```

```{r}
# Subtraction
d <- x - k
d
```

```{r}
# Division
d <- x / k
d
```

```{r}
# Square
x^2 
```

```{r}
# Pairwise comparisons
x <= k
```

### 2.2  Generates a vector of random numbers following a normal distribution using rnorm() function

```{r}
# generates 3 random number
rnorm(3)   
```

```{r}
# provide our own mean and standard deviation
v <- rnorm(8, mean=10, sd=2)
v
```

* **Some common Function on Vector in R**
* max(), min(), length(), sum().
* mean(), median(), var(), sd().
* summary()







# Matrices 
* A matrix can be created with the matrix() function.

* The arguments to this matrix() are the set of elements in the vector.

* Specify how many numbers of rows using nrow and how many numbers of columns using ncol to get the amount of rows and columns.

* Note: By default, matrices are in column-wise order.

* Create a matrix a blank matrix

```{r}
A = matrix(,nrow = 6, ncol = 4)
A
```

* creating a matrix 
```{r}
# In R programming, a 2D array is referred to as a "matrix"
m <- matrix(c(3,4,5,6), nrow = 2, ncol = 2)
m
```

```{r}
# To determines if an object is a matrix
is.matrix(m)
```

* create a matrix with strings
```{r}
# By default, matrices are in column-wise order
# to make it in row, byrow = TRUE,
m1<- matrix(c("orange", "lime","red","blue"), nrow = 2, ncol = 2)
m1
```

```{r}
#The Number of rows and columns in a Matrix
dim(m) 
```

```{r}
#The dimension of a Matrix
length(m)
```

* Acessing the Matrix Items
* The items can be accessed with the brackets [ ]. The first number, “1,” in the bracket, specifies the row position, while the second, “2,” specifies the column position.

```{r}
m <- matrix(c(3,4,5,6), nrow = 2, ncol = 2)
m
```

```{r}
# First row and second column
m[1, 2]
```

```{r}
# To access the whole row
m[1, ]
```

```{r}
# To access the whole column
m[, 2]
```

* Access More Than One Row/or Access More Than One Column
```{r}
mat <- matrix(c(1,2,3,4,5,6,7,8,9), nrow = 3, ncol = 3)
mat
```

```{r}
#Access More Than One Row
mat[c(1,3), ]
```

```{r}
#Access More Than One Column
mat[ , c(1,2)]
```

* Add Rows and Columns

* Use the rbind() function to add additional rows in a Matrix.

* Use the cbind() function to add additional columns in a Matrix

* Note: The cells in the new row must be of the same length as the existing matrix.

```{r}
#Add additional rows in a Matrix
newmatrix <- rbind(mat, c(20,30,40))
newmatrix
```

```{r}
#Add additional columns in a Matrix 
newmatrix <- cbind(mat, c(20,30,40))
newmatrix
```

```{r}
#Remove the first row and the first column
newmat <- mat[-c(1), -c(1)]
newmat
```

* combine two Matrices
```{r}
Mat1 <- matrix(c(1,2,3,4), nrow = 2, ncol = 2)
Mat2 <- matrix(c(10,20,30,40), nrow = 2, ncol = 2)
# Adding it as a rows
Mat_Combined <- rbind(Mat1, Mat2)
Mat_Combined
```

```{r}
# Adding it as a columns
Mat_Combined <- cbind(Mat1, Mat2)
Mat_Combined
```

## 3.1  Operations on Matrices in R

* Matrices Multiplication

* For matrix multiplication to be valid, the number of columns in the first matrix (mat3) must equal the number of rows in the second matrix (mat4).

* The resulting matrix will have the same number of rows as the first matrix (mat3) and the same number of columns as the second matrix (mat4).

```{r}
mat3 <- matrix(rep(5,9), ncol = 3)
mat3
```

























